preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-14 Manuscript/manuscript.Rmd
updated

**Methods** To improve the detection of domain boundaries, we developed preciseTAD, a random forest-based framework that leverages high-resolution genome annotation data coupled with density-based clustering and scalable partitioning techniques. We investigated several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) to address class imbalance. We benchmarked our method against Arrowhead domain caller and a novel chromatin loop prediction algorithm, Peakachu, on multiple cell lines.

@@

Several methods have been proposed to identify the genomic coordinates of TAD boundaries (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers, referred hereafter as domain callers, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic regions (bins) used to segment the linear genome [??? Is this the right statement and reference? as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison] ???]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among domain callers is that they disregard prior knowledge about functional genomic annotations associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been identified as components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Additionally, distinct patterns of histone modifications have also been shown to be present at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These genomic annotations are obtained using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). The resolution of ChIP-seq experiments is typically on the order of tens to hundreths bases [@mendoza2009genome], well below tens of kilobases resolution of Hi-C data (750 bp is the highest resolution of Hi-C data to date [@bonev2017multiscale]). Therefore, leveraging precisely mapped genomic annotations in a supervised machine learning framework provides the possibility of more precisely predicting the locations of TAD (or loop) boundaries.

Here we present *preciseTAD*, an R package implementing optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes random forest (RF) algorithm trained on chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data. We introduce a systematic pipeline for building optimal boundary prediction classifier. We found that spatial associations (distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines. This improved precision in TAD and loop boundary location can provide insight into the role of genomic regulators on the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).

@@

Publicly available TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools, respectively, were used for training. Arrowhead TAD boundaries were called from primary Hi-C data for the GM12878 and K562 cell lines [@rao20143d]. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site (Supplementary Table 1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 3).

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. A *holdout chromosome* technique was used for model evaluation on each autosomal chromosome. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. $p$, the number of annotations, together with the response vector **Y**, defined column-wise dimension of the matrix $A$. Re-sampling was then performed on $A$ and an RF classifier was trained using 3-fold cross-validation to tune for the number of annotations to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

To investigate whether we could alleviate resolution limitations of conventional domain calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base resolution. To demonstrate, we implemented an optimally trained RF model using the framework described above. That is, a random forest model was built on the optimal combination of predictor type (predictor space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements) to maximaly accurately predict boundaries at _Hi-C resolution level_.

To precisely idenify boundary locations, we first construct a _base-level resolution_ predictor space for a particular chromosome **i**. We then apply our optimally trained RF model on the base-level predictor space to extract the probability vector, $p_{n_{i}}$, denoting each base's probability of being a boundary point. Here, $n_{i}$ represents the length of chromosome **i**. Next, DBSCAN (version 1.1-5) was applied to the matrix [??? vector?] of pairwise genomic distances between bases with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close bases with similar predictor profiles (i.e., similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The *eps* parameter was assigned as the respective data resolution. The heuristic being that density-reachable bases with genomic distances less than resolution should occupy the same designated cluster. The corresponding *MinPts* parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive bases identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (**Clustering Large Applications**) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base coordinate within each clustered PTBR.

@@

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for precise prediction of boundaries of Topologically Associated Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We use published Arrowhead-called TAD boundaries [@durand2016juicer] and Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Supplementary Table 1). We developed *preciseTAD*, an R package which leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base resolution from cell type-specific genome annotation data.

In our ML framework boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 2). The number of non-boundary regions highly outnumbered boundary regions. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 cell line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 3). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These include signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods). A ($log_2 + 1$)-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1).

@@

![**A diagram of the machine learning framework used to build TAD/loop boundary region prediction models.** Step 1 defines the data matrix A, composed of N genomic bins for each chromosome, except a holdout chromosome *i*, *p* genomic annotations, and the response vector **Y** (for *p*+1 columns). Resolution-specific data (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the n-1 autosomal chromosomes. Step 2 is to reserve the predictor-response matrix for the holdout chromosome *i* as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the mtry parameter. Finally, step 5 is to validate the model on the separate test data composed of the binned data from the holdout chromosome *i* and to evaluate model performance using balanced accuracy (BA).](Figures/figure2.png)

@@

We hypothesized that the type of genomic annotations may also affect the predictive performance. Using the established optimal settings (RUS, Distance, 5 kb bins), we built separate random forest models using BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both the GM12878 and K562.

@@

We sought to further optimize our boundary region prediction models by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations vary for each chromosome. To determine transcription factors most important for boundary prediction, we clustered the predictive importances (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were consistently predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained on a bin-level space is used to predict the probability of each base being a boundary. Our method, referred to as *preciseTAD*, uses density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location at base resolution (see Methods).

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7,846 TAD and 15,838 chromatin loop boundaries on GM12878, as well as 8,883 TAD and 11,016 chromatin loop boundaries on K562. Noticeably, we reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 2). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that often do not correspond to intra-chromosomal contacts featured on a contact matrix. <!--Likewise, Arrowhead often produces nested and overlapped TADs that do not have demarcated boundaries, further exacerbating this inflation. However, under the framework of *preciseTAD*, any two domains are inherently conjoined by only one predicted boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] -->Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 2). This can be contributed to the fact that *preciseTAD* leverages four known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting Peakachu which considers only CTCF sites as true positives. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant, in contrast to boundaries detected from Hi-C data alone.

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps indicated that our predicted boundaries were more frequently co-localized with CTCF sites than either Arrowhead or Peakachu boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more co-localized with other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 5-6). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were statistically significantly closer to the same cell line-specific top ranked TFBS (p<0.001 and p<0.001 versus Arrowhead and Peakachu respectively) (Figure 5D, Supplementary Figure 7). These results indicate that *preciseTAD*-predicted boundaries better reflect the known bioloby of boundary formation.

@@

<!--Additionally, the median Jaccard index (MJI) was found to be 0.472 (IQR=0.441-0.503; p=???) and 0.647 (IQR=0.646-0.647; p=???) for *preciseTAD* predicted TAD and loop boundaries (Supplementary Table 5). Comparatively, only 25% (MJI=0.284; IQR=0.268-0.301) and 49% (MJI=0.509; IQR=0.503-0.515) of called/predicted Arrowhead and Peakachu boundaries were conserved (Figure 6A, Supplementary Table 5). These results demonstrate that boundaries predicted using genomic annotations (*preciseTAD*) are better conserved across cell lines.-->

<!--We also inspected the effect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than Arrowhead boundaries, thus implying stronger invariance toward resolution (Figure 5E).-->

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on ground truth TAD/loop boundaries obtained from a TAD/loop calling tools and genomic annotations as the predictor space. We benchmarked *preciseTAD* against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We applied *preciseTAD* on cell line-specific boundaries and annotation data from both GM12878 and K562. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and re-sampling techniques. We show that genomic distance between high-resolution boundary regions and transcription factor binding sites, coupled with random under-sampling results in best model performance. We demonstrate that binding of four transcription factors (SMC3, Rad21, CTCF, Znf143) are sufficient for accurate TAD and loop boundary predictions. Applied to each annotated base, the model predicts probability for each base to be a boundary, and, coupled with spatial clustering (DBSCAN) and scalable partition around medoids clustering (CLARA), allows for the detection of boundaries at the base-level resolution. *preciseTAD* is implemented as an R package, and allows for precise prediction of boundaries for any cell type for which the corresponding genomic annotations are available.

>>>>

**Methods** To improve the detection of domain boundaries, we developed *preciseTAD*, a random forest-based framework that leverages high-resolution genome annotation data coupled with density-based clustering and scalable partitioning techniques. We investigated several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) to address class imbalance. We benchmarked our method against the Arrowhead domain caller and a novel chromatin loop prediction algorithm, Peakachu, on multiple cell lines.

@@

Several methods have been proposed to identify the genomic coordinates of TAD boundaries (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers, referred hereafter as domain callers, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic regions (bins) used to segment the linear genome. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among domain callers is that they disregard prior knowledge about functional genomic annotations associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been identified as components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Additionally, distinct patterns of histone modifications have also been shown to be present at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These genomic annotations are obtained using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). The resolution of ChIP-seq experiments is typically on the order of tens to hundreds of bases [@mendoza2009genome], well below the resolution of Hi-C data (tens of kilobases; 750 bp is the highest resolution of Hi-C data to date [@bonev2017multiscale]). Therefore, leveraging precisely mapped genomic annotations in a supervised machine learning framework provides the possibility of more precisely predicting the locations of TAD (or loop) boundaries.

Here we present *preciseTAD*, an R package for implementing an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data. We introduce a systematic pipeline for building optimal boundary prediction classifier. We found that spatial associations (distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines. This improved precision in TAD and loop boundary location can provide insight into the role of genomic regulators on the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).

@@

Publicly available TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools, respectively, were used for training. Arrowhead TAD boundaries were called from primary Hi-C data for the GM12878 and K562 cell lines [@rao20143d]. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site (Supplementary Table 1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2).

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. A *holdout chromosome* technique was used for model evaluation on each autosomal chromosome. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. $p$, the number of annotations, together with the response vector **Y**, defined the column-wise dimension of the matrix $A$. Re-sampling was then performed on $A$ and an RF classifier was trained using 3-fold cross-validation to tune for the number of annotations to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

To investigate whether we could alleviate resolution limitations of conventional domain calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base resolution. To demonstrate, we implemented an optimally trained RF model using the framework described above. That is, a random forest model was built on the optimal combination of predictor type (predictor space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements) to maximally accurately predict boundaries at base _Hi-C resolution level_.

To precisely identify boundary locations, we first construct a _base-level resolution_ predictor space for a particular chromosome *i*. We then apply our optimally trained RF model on the base-level predictor space to extract the probability vector, $p_{n_{i}}$, denoting each base's probability of being a boundary point. Here, $n_{i}$ represents the length of chromosome *i*. A threshold (*t*) is specified on $p_{n_{i}}$, such that base coordinates with $p_{n_{i}} \ge t$ are extracted. Next, density-based spatial clustering of applications with noise (DBSCAN; version 1.1-5) is applied to the matrix of pairwise genomic distances between bases with $p_{n_{i}} \ge t$. The resulting clusters of highly predictive bases identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (**Clustering Large Applications**) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base coordinate within each clustered PTBR.

The DBSCAN algorithm takes two parameters, *MinPts* and *eps* ($\epsilon$). The *MinPts* parameter was set to the recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. To decide on the optimal value of *t* and $\epsilon$ in *preciseTAD*, we considered the normalized enrichment of flanked predicted boundaries, defined as

$$N_{e} = \dfrac{1}{b} \Sigma_{i=1}^{n}e_{i}$$
where $e_{i}$ is the number ChIP-seq peak regions that overlap with the $i^{th}$ predicted boundary, $b$, flanked by resolution $r$. We evaluated $N_{e}$ for combinations of $t=\{0.95, 0.975, 0.99, 1.0 \}$ and $\epsilon = \{5000, 10000, 15000, 20000, 25000 \}$. The heuristic of $\epsilon$ is that density-reachable bases with genomic distances less than $\epsilon$ should occupy the same designated cluster.

@@

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for precise prediction of boundaries of Topologically Associated Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We use published Arrowhead-called TAD boundaries [@durand2016juicer] and Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Supplementary Table 1). We next developed *preciseTAD*, an R package which leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base resolution from cell type-specific genome annotation data.

In our ML framework boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 cell line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 2). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods). A ($log_2 + 1$)-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1).

@@

![**A diagram of the machine learning framework used to build TAD/loop boundary region prediction models.** Step 1 defines the dimensions of the data matrix $A_{N \times (p+1)}$ used for training and testing by establishing a resolution, $r$, at which to bin the linear genome into $N$ genomic bins for each chromosome, except a holdout chromosome *i*. The response vector **Y** is defined as a boundary region ($Y=1$) if it overlaps with a genomic bin (else $Y=0$) A feature engineering procedure is selected to define the feature space of $A$ composed of $p$ genomic annotations. Step 2 is to reserve the predictor-response matrix for the holdout chromosome *i* as the test data. Step 3 is to apply a re-sampling technique to the training data to address class imbalance. Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the mtry parameter. Finally, step 5 is to validate the model on the separate test data composed of the binned data from the holdout chromosome *i* and to evaluate model performance using balanced accuracy (BA).](Figures/figure2.png)

@@

We hypothesized that the type of genomic annotations may also affect predictive performance. Using the established optimal settings (RUS, Distance, 5 kb bins), we built separate random forest models using BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both GM12878 and K562.

@@

We sought to further optimize our boundary region prediction models by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations varied for each set of chromosomes. To determine transcription factors most important for boundary prediction, we clustered the predictive importances (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were consistently predictive of TAD and loop boundaries (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. [???update]

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space was built on a base-resolution level, and the model trained on a bin-level space was used to predict the probability of each base being a boundary. Our method, referred to as *preciseTAD*, uses density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location at base resolution (see Methods).

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7,846 TAD and 15,838 chromatin loop boundaries on GM12878, as well as 8,883 TAD and 11,016 chromatin loop boundaries on K562. Noticeably, we reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 3). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that often do not correspond to intra-chromosomal contacts featured on a contact matrix. Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 3). This can be contributed to the fact that *preciseTAD* leverages four known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting Peakachu which considers only CTCF sites as true positives. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant, in contrast to boundaries detected from Hi-C data alone.

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps indicated that our predicted boundaries were more frequently co-localized with CTCF sites than either Arrowhead or Peakachu boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more co-localized with other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 5-6). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were statistically significantly closer to the same cell line-specific top ranked TFBS (p<0.001 and p<0.001 versus Arrowhead and Peakachu respectively) (Figure 5D, Supplementary Figure 7). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

@@

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on ground truth TAD/loop boundaries obtained from TAD/loop calling tools and genomic annotations as the predictor space. We benchmarked *preciseTAD* against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We applied *preciseTAD* on cell line-specific boundaries and annotation data from both GM12878 and K562. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and re-sampling techniques. We show that genomic distance between high-resolution boundary regions and transcription factor binding sites, coupled with random under-sampling results in best model performance. We demonstrate that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) are sufficient for accurate TAD and loop boundary predictions. Applied to each annotated base, the model predicts probability for each base to be a boundary, and, coupled with spatial clustering (DBSCAN) and scalable partition around medoids clustering (CLARA), allows for the detection of boundaries at the base-level resolution. *preciseTAD* is implemented as an R package, and allows for precise prediction of boundaries for any cell type for which the corresponding genomic annotations are available.
